### 2012-09-16 paper.tex
minor sentence edit
second set of changes

In professional tennis, a player may `retire hurt' from a match at any time should they feel they are unable to complete the match due to injury or illness, or that it is unwise to continue in case they aggravate their condition.  The match is consequently awarded to the opponent regardless of the current match state.  A \textit{walkover} occurs when a player withdraws from a match before it has begun.  In-play injuries are common occurrences in tennis as a whole.  Between 2000 and 2009, there was a retirement during approximately 3.9\% of Grand Slam men's singles matches\footnote{http://www.tennis.ukf.net/stats15.htm}.

Betting companies take different approaches when dealing with the issue of player retirement.  Typically, they fall into one of four categories (with regards to Match Odds markets)\footnote{http://rebelbetting.com/faq/tennis-rules}:

@@

Markets offered on the outcomes of individual games or sets are always rendered void unless the result has been unconditionally determined.

@@

UK-based Betfair, the world's first betting exchange, falls into \textit{Category 2} of the tennis betting retirement payout policies with respect to its in-play Match Odds market.  Betfair's Set Betting market is always voided in the event of a retirement since if the match is not finished, we do not have a final score.

@@

Compare with Figure \ref{murrayberrer} which displays the evolution of implied match-winning probabilities for Andy Murray when he played Michael Berrer in the French Open 2011 Men's Third Round.  In particular, observe where the Match Odds probability of Murray winning the match suddenly drops \textit{almost 60\%}.  This phenomenon in the odds data occurred during a \textit{single point in the match}.  There is no possible event that could have caused this huge swing in the market so quickly other than the \textit{injury} that was suffered by Andy Murray.  We quote from the \textit{BBC Sport}\footnote{http://www.bbc.co.uk/sport} live text commentary of the match during this point:

\begin{itemize}
	\item "Big, big trouble for Andy Murray, who has gone over on his right ankle and looks in real pain. Not sure whether he will be able to continue. Unbelievable."
	\item "We are going to have a medical time-out while Murray has treatment. He slipped as he ran in to put away a forehand, and the replays are not very pleasant to watch."
\end{itemize}

@@

We can observe the evolution of the Betfair Set Betting and Match Odds markets in order to help us \textit{quantify} a player's risk of retirement.  For example, we can see from the Murray vs Berrer match when the injury occurred that the Match Odds implied probability dropped sharply but the Set Betting implied probability remained relatively stable.  If a player becomes injured and concedes the match at any point, traditional bookmakers will usually cancel bets in all markets (they fall into Category 4).  Betfair will refund money on final score bets (Set Betting) but will still pay out on bets to win (Match Odds) as long as at least one set has been played (since the match has a winner but no final score).  In this way, the Betfair Set Betting market \textit{simulates} a Match Odds market that \textbf{ignores risk of retirement} whereas the actual Betfair Match Odds market \textbf{takes into account retirement risk}, given one set has been played.  Consequently, we see discrepancies between the match and final score odds in such scenarios (possibly more so beyond the first set) as we do in the Murray match.

The theory is that the probability of a player retiring \textit{at some point during the remainder of the match} should be somehow encapsulated within the difference between the two markets, i.e. the combined opinions of all the traders betting in those markets (the wisdom of the crowd).  All the complex factors which lead to a player deciding to retire such as the state of the match, the seriousness of the injury, the importance of the match, even their past history of retirements, should be \textit{some function} of this gap in the odds.

@@

There have been many past works on the topic of modelling tennis.  \cite{omalley} presents what are considered the \textit{tennis formulae}.  The tennis formulae are a hierarchical series of equations that compute the probability a given player will win a tennis match given the probabilities that the given player and his/her opponent will win any of their service points.  Combined together are individual formulae for the probabilities of winning games, sets, and tiebreaks.

\cite{servefirst} more comprehensively explore the use of recurrence relations to model tennis, utilising them to calculate probabilities of winning tournaments and also proving explicitly that the probability of winning a set or match does not depend on which player serves first.  \cite{excel} experiment with the same idea in Microsoft Excel.  They investigate using six parameters rather than just the two point-winning probabilities taking into account service faults.  Consequently, for each player they input the probability of a successful first serve, the probability of winning a point on first serve, and the probability of winning a point on second serve.  \cite{willyk} also designed a low-level model of tennis points, taking into account such events as service faults and service aces, in order to predict the outcome of matches.  They tailor their input probabilities to the specific returning capabilities of the given opponent (not an average player), estimated using historical statistics.  They find that their model was able to successfully predict 67\% of outcomes from a sample of 1839 ATP matches played during 2011.  Since we are already adding further complexity with the probability of retirement, we shall concentrate on extending a simpler, two parameter version of a tennis model.  \cite{momentum} continued with an investigation into player momentum in tennis matches by slightly perturbing point-winning probability depending on how much the given player is leading or trailing the match by (essentially introducing a dependency between points).

@@

@@

The granularity of the model is point-level as this is compatible with the structure of standard tennis models.  When a tennis player plays a point, he or she puts great strain on their body.  This strain naturally leads to a chance of an injury occurring.  For the vast majority of points played, the strain is perfectly manageable and does not lead to injury.  For example, Grand Slam winners will hit over 1000 serves during the course of the tournament without issue.  Occasionally however, the body fails to cope with the strain, or the strain is for some reason much more acute than normal (e.g. remember Andy Murray twisting his ankle), and an injury occurs.  When a player does get injured during a match, it does not always end in retirement.  Players often soldier on at least for a few points and may even recover from the injury as the match progresses.

Using this intuition, we present an elegant model for the probability a player will retire \textit{on a given point} in a tennis match:

@@

\begin{figure}[H]\Large
	\begin{center}
		\begin{tabular}{|C{3.5cm}|C{2.5cm}|C{0.7cm}|C{0.7cm}|}
			\multicolumn{1}{l}{\small{0}} & \multicolumn{1}{r}{} & \multicolumn{1}{r}{} & \multicolumn{1}{r}{\small{1}} \\
			\hline
			$p$ & $q$ & $r_A$ & $r_B$ \\ \hline
		\end{tabular}
	\end{center}
	\caption{The four possible outcomes of a point in our modified tennis match simulator}
	\label{bins}
\end{figure}

@@

We have introduced three unknowns into our simulator; $c$ is the success parameter of the Bernoulli distribution (per point injury probability), $X$, $\lambda$ is the rate parameter of the injury magnitude truncated exponential distribution, $Y$, and $\rho$ is the injury recovery factor.

Figure \ref{variabletable} displays a table describing the majority of the variables we have introduced thus far.  An appropriate parameterisation of the model would be values for the unknown input variables, $c$, $\lambda$, and $\rho$, that generate $R_A \approx R_B \approx 1.95\%$ when input into our modified simulator at the start of a 5-set match.  We believe these match-level retirement risks are justified as it corresponds to the 3.9\% chance of such a match ending in retirement as seen in men's Grand Slam singles matches between 2000 and 2009.  Given \cite{dominance} find that the average point-winning probability on serve for a top-level professional tennis player is 0.645 for men and 0.560 for women, we choose 0.6 as a reasonable value for $P_A$ and $P_B$ which denotes an even match.  The goal now is to \textit{fit} the model to accurately reflect real-world events.

@@

Vitally important to the success of the Nelder-Mead method is the choice of objective function to minimise.  In our case, our modified simulator is essentially the function, but we must still define what it means to minimise it.  This is the main way we mould the algorithm to solve our problem.  We want the difference between the probability Player A wins the match normally given no risk of retirement and the probability Player A wins the match normally with risk of retirement to be $0.0195$, which is the probability Player A wins the match via Player B retiring (and similarly for Player B).  Consequently, we aim to minimise the below expression:

@@

Specifically, at the minimum and since the match is even, we would have:

@@

Our model appears to produce reasonable results from the beginning of a match but this is only because of our assumption that players begin matches with no risk of retiring ($r_0^A = r_0^B = 0$).  Our model must also be applicable for any given match state, including the situation where an injury has recently occurred but the affected player is continuing to play.  At this stage, we know the gap in the odds for each player ($G_A$ and $G_B$), but how can we go \textit{backwards} from this and calculate the retirement risk for the \textit{current point}, $r_n^A$ and $r_n^B$ (which will naturally be a lot smaller)?  Well once again, we can use the Nelder-Mead method to approximate these values for each point in the match.  Given this information, our simulator will be able to calculate the match-level retirement risk for each player at each point ($R_n^A$ and $R_n^B$).

@@

Up until this point, we have assumed that we will be able to manually input the point-winning probabilities on serve of both players in the match.  Due to \cite{dominance}, we know that the average point-winning probability on serve for a top-level professional tennis player, $\gamma$, is 0.645 for men and 0.560 for women.  We also know, according to \cite{marek}, that the important thing in determining the winner of a match is the \textit{difference}, $\delta$, between the point-winning probabilities of each player and not the absolute values.  Consequently, if we can find what $\delta$ \textit{should} be, we can choose appropriate values for $P_A$ and $P_B$ by using the constraint that their average equals $\gamma$.  We must find both; we cannot, for example, fix $P_A$ to $\gamma$ and linearly search for $P_B$, since the model would cease to be symmetric for both players.  We know the current implied match-winning probability of the Set Betting (no retirement risk) market for both players and so we can do a simple binary search to find a value for $\delta$ (and therefore values for $P_A$ and $P_B$) for which O'Malley's tennis formulae generate the same results as the \textit{no risk} market.  Marek also states that for one to be able to express match-winning probability as a function of $\delta$, one must constrain $-0.1 \leq \delta \leq 0.1$.  Since the market's opinion of each player's point-winning probability changes throughout the match, we recalculate $P_A$ and $P_B$ on every point.  Note that this does not invalidate the assumption that points are iid as no dependency between points is introduced.

@@

We have created a new model for retirement risk in tennis matches and implemented it into a tennis match simulator by incorporating extra parameters approximated using real-world averages and betting odds data as well as additional match outcomes compared to standard tennis models.  This is the world's first attempt to create such a model.  Testing the model in a totally artificial environment, we found it was able to mimic the expected patterns of in-play betting markets with different retirement payout policies for matches with varying injury scenarios.

>>>>

In professional tennis, a player may `retire hurt' from a match at any time should they feel they are unable to complete the match due to injury or illness, or that it is unwise to continue in case they aggravate their condition.  The match is consequently awarded to the opponent regardless of the current match state.  A \textit{walkover} occurs when a player withdraws from a match before it has begun.  In-play injuries are common occurrences in tennis as a whole.  Between 2000 and 2009, there was a retirement during approximately 3.9\% of Grand Slam men's singles matches\footnote{http://www.tennis.ukf.net/stats15.htm}.  Betting companies take different approaches when dealing with the issue of player retirement.  Typically, they fall into one of four categories (with regards to Match Odds markets)\footnote{http://rebelbetting.com/faq/tennis-rules}:

@@

Markets offered on the outcomes of individual games or sets are always rendered void unless the result has been unconditionally determined.  UK-based Betfair, the world's first betting exchange, falls into \textit{Category 2} of the tennis betting retirement payout policies with respect to its in-play Match Odds market.  Betfair's Set Betting market is always voided in the event of a retirement since if the match is not finished, we do not have a final score.

@@

@@

Compare with Figure \ref{murrayberrer} which displays the evolution of implied match-winning probabilities for Andy Murray when he played Michael Berrer in the French Open 2011 Men's Third Round.  In particular, observe where the Match Odds probability of Murray winning the match suddenly drops \textit{almost 60\%}.  This phenomenon in the odds data occurred during a \textit{single point in the match}.  Although there are a few other anomalous events could have caused this huge and rapid swing in the market such as a disqualification, the cause could only have been the \textit{injury} that was suffered by Andy Murray.  We quote from the \textit{BBC Sport}\footnote{http://www.bbc.co.uk/sport} live text commentary of the match during this point:

\begin{itemize}
	\item ``Big, big trouble for Andy Murray, who has gone over on his right ankle and looks in real pain. Not sure whether he will be able to continue. Unbelievable.''
	\item ``We are going to have a medical time-out while Murray has treatment. He slipped as he ran in to put away a forehand, and the replays are not very pleasant to watch.''
\end{itemize}

@@

We can observe the evolution of the Betfair Set Betting and Match Odds markets in order to help us \textit{quantify} a player's risk of retirement.  For example, we can see from the Murray vs Berrer match when the injury occurred that the Match Odds implied probability dropped sharply but the Set Betting implied probability remained relatively stable.  If a player becomes injured and concedes the match at any point, traditional bookmakers will usually cancel bets in all markets (they fall into Category 4).  Betfair will refund money on final score bets (Set Betting) but will still pay out on bets to win (Match Odds) as long as at least one set has been played (since the match has a winner but no final score).  In this way, the Betfair Set Betting market \textit{simulates} a Match Odds market that \textbf{ignores risk of retirement} whereas the actual Betfair Match Odds market \textbf{takes into account retirement risk}, given one set has been played.  Consequently, we see discrepancies between the match and final score odds in such scenarios (possibly more so beyond the first set) as we do in the Murray match.  
The theory is that the probability of a player retiring \textit{at some point during the remainder of the match} should be somehow encapsulated within the difference between the two markets, i.e. the combined opinions of all the traders betting in those markets (the wisdom of the crowd).  All the complex factors which lead to a player deciding to retire such as the state of the match, the seriousness of the injury, the importance of the match, even their past history of retirements, should be \textit{some function} of this gap in the odds.

@@

There have been many past works on the topic of modelling tennis.  \cite{omalley} presents what are considered the \textit{tennis formulae}.  The tennis formulae are a hierarchical series of equations that compute the probability a given player will win a tennis match given the probabilities that the given player and his/her opponent will win any of their service points.  Combined together are individual formulae for the probabilities of winning games, sets, and tiebreaks.  
\cite{servefirst} more comprehensively explore the use of recurrence relations to model tennis, utilising them to calculate probabilities of winning tournaments and also proving explicitly that the probability of winning a set or match does not depend on which player serves first.  \cite{excel} experiment with the same idea in Microsoft Excel.  They investigate using six parameters rather than just the two point-winning probabilities taking into account service faults.  Consequently, for each player they input the probability of a successful first serve, the probability of winning a point on first serve, and the probability of winning a point on second serve.  \cite{willyk} also designed a low-level model of tennis points, taking into account such events as service faults and service aces, in order to predict the outcome of matches.  They tailor their input probabilities to the specific returning capabilities of the given opponent (not an average player), estimated using historical statistics.  They find that their model was able to successfully predict 67\% of outcomes from a sample of 1839 ATP matches played during 2011.  Since we are already adding further complexity with the probability of retirement, we shall concentrate on extending a simpler, two parameter version of a tennis model.  \cite{momentum} continued with an investigation into player momentum in tennis matches by slightly perturbing point-winning probability depending on how much the given player is leading or trailing the match by (essentially introducing a dependency between points).

@@

\newpage

@@

The granularity of the model is point-level as this is compatible with the structure of standard tennis models.  When a tennis player plays a point, he or she potentially puts great strain on their body.  This strain naturally leads to a chance of an injury occurring.  For the vast majority of points played, the strain is perfectly manageable and does not lead to injury.  For example, Grand Slam winners will hit over 1000 serves during the course of the tournament without issue.  Occasionally however, the body fails to cope with the strain, or the strain is for some reason much more acute than normal (e.g.\ remember Andy Murray twisting his ankle), and an injury occurs.  When a player does get injured during a match, it does not always end in retirement.  Players often soldier on at least for a few points and may even recover from the injury as the match progresses.  
Using this intuition, we present an elegant model for the probability a player will retire \textit{on a given point} in a tennis match:

@@

\begin{figure}[H]\Large
	\begin{center}
		\begin{tabular}{|C{3.5cm}|C{2.5cm}|C{0.7cm}|C{0.7cm}|}
			\multicolumn{1}{l}{\small{0}} & \multicolumn{1}{r}{} & \multicolumn{1}{r}{} & \multicolumn{1}{r}{\small{1}} \\
			\hline
			$P_A$ & $1 - P_A$ & $r_A$ & $r_B$ \\ \hline
		\end{tabular}
	\end{center}
	\caption{The four possible outcomes of a point in our modified tennis match simulator where Player A is serving}
	\label{bins}
\end{figure}

@@

We have introduced three unknowns into our simulator; $c$ is the success parameter of the Bernoulli distribution (per point injury probability) corresponding to random variable $X$, $\lambda$ is the rate parameter of the injury magnitude truncated exponential distribution corresponding to random variable $Y$, and $\rho$ is the injury recovery factor.

Figure \ref{variabletable} displays a table describing the majority of the variables we have introduced thus far.  An appropriate parameterisation of the model would be values for the unknown input variables, $c$, $\lambda$, and $\rho$, that generate $R_A \approx R_B \approx 1.95\%$ when input into our modified simulator at the start of a 5-set match.  We believe these match-level retirement risks are justified as it corresponds to the 3.9\% chance of such a match ending in retirement as seen in men's Grand Slam singles matches between 2000 and 2009.  Given \cite{dominance} find that the average point-winning probability on serve for a top-level professional tennis player is 0.645 for men and 0.560 for women, we choose 0.6 as a reasonable value for $P_A$ and $P_B$ which denotes an even match.  The goal now is to \textit{fit} the model by adjusting the parameter values to accurately reflect real-world events.

@@

Vitally important to the success of the Nelder-Mead method is the choice of objective function to minimise.  In our case, our modified simulator is essentially the function, but we must still define what it means to minimise it.  This is the main way we mould the algorithm to solve our problem.  We want the difference between the probability Player A wins the match normally given no risk of retirement and the probability Player A wins the match normally with risk of retirement to initially be $0.0195$, which is the probability at the beginning of the match that Player A wins the match via Player B retiring (and similarly for Player B).  Consequently, we aim to minimise the below expression:

@@

More specifically, at the minimum and since the contest is evenly matched, we would have:

@@

Our model appears to produce reasonable results from the beginning of a match given our assumption that players begin matches with no risk of retiring ($r_0^A = r_0^B = 0$).  Our model must also be applicable for any given match state, including the situation where an injury has recently occurred but the affected player is continuing to play.  At this stage, we know the gap in the odds for each player ($G_A$ and $G_B$), but how can we go \textit{backwards} from this and calculate the retirement risk for the \textit{current point}, $r_n^A$ and $r_n^B$ (which will naturally be a lot smaller)?  Well once again, we can use the Nelder-Mead method to approximate these values for each point in the match.  Given this information, our simulator will be able to calculate the match-level retirement risk for each player at each point ($R_n^A$ and $R_n^B$).

@@

Up until this point, we have assumed that we will be able to manually input the point-winning probabilities on serve of both players in the match.  Due to \cite{dominance}, we know that the average point-winning probability on serve for a top-level professional tennis player, $\gamma$, is 0.645 for men and 0.560 for women.  We also have the insight, arising from \cite{omalley} and described by \cite{marek}, that the important thing in determining the winner of a match is the \textit{difference}, $\delta$, between the point-winning probabilities of each player and not the absolute values.  Consequently, if we can find what $\delta$ \textit{should} be, we can choose appropriate values for $P_A$ and $P_B$ by using the constraint that their average equals $\gamma$.  We must find both; we cannot, for example, fix $P_A$ to $\gamma$ and linearly search for $P_B$, since the model would cease to be symmetric for both players.  We know the current implied match-winning probability of the Set Betting (no retirement risk) market for both players and so we can do a simple binary search to find a value for $\delta$ (and therefore values for $P_A$ and $P_B$) for which O'Malley's tennis formulae generate the same results as the \textit{no risk} market.  Marek also states that for one to be able to express match-winning probability as a function of $\delta$, one must constrain $-0.1 \leq \delta \leq 0.1$.  Since the market's opinion of each player's point-winning probability changes throughout the match, we recalculate $P_A$ and $P_B$ on every point.  Note that this does not invalidate the assumption that points are iid as no dependency between points is introduced.

@@

We have created a new model for retirement risk in tennis matches and implemented it into a tennis match simulator by incorporating extra parameters approximated using real-world averages and betting odds data as well as additional match outcomes compared to standard tennis models.  To the best of our knowledge, this is the world's first published attempt to create such a model.  Testing the model in a totally artificial environment, we found it was able to mimic the expected patterns of in-play betting markets with different retirement payout policies for matches with varying injury scenarios.
